String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-10-19 2023-1-19/sections/algorithms.tex
remove comments

%!TEX root = ../main.tex
%\documentclass{standalone}
%\begin{document}

@@

% \vspace{-2mm}
\begin{definition}[Nonemptiness of CEFA w.r.t. an LIA formula
% \footnote{The decision problem was called satisfiability of an LIA formula w.r.t. CEFA in \cite{atva2020}.}
, abbreviated as $\cefadec$]
Suppose that $x_1, \cdots, x_n$ are string variables, $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are nonempty sets of CEFA over the alphabet $\Sigma$ with $\Lambda_{x_i} = \{\aut_{i, 1}, \cdots, \aut_{i, l_i}\}$ for every $i \in [n]$, and the sets of registers $R_{\aut_{1, 1}}, \cdots, R_{\aut_{1, l_1}}, \cdots, R_{\aut_{n, 1}}, \cdots, R_{\aut_{n, l_n}}$ are mutually disjoint, moreover, $\varphi$ is an LIA formula whose free variables are from $ \bigcup \limits_{i \in [n], j \in [l_i]} R_{\aut_{i, j}}$. Then  the CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are said to be nonempty w.r.t. $\varphi$ if there are assignments $\theta: \{x_1, \cdots, x_n\} \rightarrow \Sigma^*$ and vectors $\myvec{v_{i,j}}$
%$\myvec{v_{1,1}}, \cdots, \myvec{v_{1, l_1}}, \cdots, \myvec{v_{n, 1}}, \cdots, \myvec{v_{n, l_n}}$ 
such that $(\theta(x_i); \myvec{v_{i,j}}) \in \Lang(\aut_{i, j})$ and  $\varphi[(\myvec{v_{i,j}}/R_{\aut_{i,j}})]$ is true, for every $i \in [n], j \in [l_i]$.
% \in \aut_{i,1}(\eta(x_1)), \cdots, \myvec{v_{i, l_i}} \in \aut_{i,l_i}(\eta(x_i))$ for every $i \in [n]$  
%\cdots, \myvec{v_{n,1}} \in \aut_{n,1}(w_n), \cdots, \myvec{v_{n, l_n}} \in \aut_{n, l_n}(w_n)$ 
% $\varphi[(\myvec{v_{i,j}}/R_{\aut_{i,j}})_{i \in [n], j \in [l_i]}]$ is true. 
%$\varphi[\myvec{v_{1,1}}/R_{\aut_{1,1}}, \cdots, \myvec{v_{n, l_n}}/R_{\aut_{n, l_n}}]$ is true. 
\end{definition}
% \vspace{-4mm}
\begin{proposition}\label{prop-cefadec}
$\cefadec$ is PSPACE-complete. 
\end{proposition}
\begin{proof}
  The lower bound follows from the fact that the nonemptiness problem of the intersection of NFAs is PSPACE-complete \cite{pspace_int}. For the upper bound, \cite{atva2020} showed that $\cefadec$ is in PSPACE by reducing it to monotonic counter machines.
\end{proof}
Note that the decision procedure in \cite{atva2020} was only used to prove the upper bound in Proposition~\ref{prop-cefadec} and not implemented as a matter of fact. Instead, the symbolic model checker nuXmv was used to solve $\cefadec$. We do not rely on nuXmv in this work and shall propose a new algorithm for solving $\cefadec$ in Section~\ref{subsec:cefadec}.

%As mentioned in Section \ref{sec:overview}, the main idea of our approach is to model the counting operators symbolically by registers instead of unfolding them explicitly. Additionally, we use registers to represent string lengths. After that, the satisfiability of string constraints involves regex-counting, and string-length is reduced to the satisfiability of LIA, which off-the-shelf SMT solvers can solve. We also propose techniques to reduce automata sizes and utilize under approximations to enhance performance.
\vspace{-2mm}
\subsection{From $\reclsat$ to $\cefadec$} \label{subsec:regex2cefa}
\vspace{-1mm}

@@

Let $e$ be a register-representable regular expression over $\Sigma$. By the following procedure, we will construct a CEFA out of $e$, denoted by $\aut_e$. 
\vspace{-1mm}
\begin{enumerate}
%
\item For each sub-expression $(e')^{\{m,n\}}$ with $m \le n$ (resp. $(e')^{\{m, \infty\}}$) of $e$, we construct a CEFA $\aut_{(e')^{\{m,n\}}}$ (resp. $\aut_{(e')^{\{m, \infty\}}}$) as follows. Let $\aut_{e'} = (Q, \Sigma, \delta, I, F)$. Same as the construction for Kleene star except for the new register and updates on transitions, $\aut_{(e')^{\{m,n\}}} = ((r'), Q', \Sigma, \delta'', I', F', \alpha')$, where $r'$ is a new register, 
%
\vspace{-3mm}
$$
\begin{array}{l c l}
\delta'' & = & \{(q, a, q', (0)) \mid (q, a, q') \in \delta\}\  \cup \\
& & \{(q_0, a, q', (1))  \mid \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\}\ \cup \\
& & \{(q, a, q', (1)) \mid q \in F, \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\},
\end{array}
$$ 
%
moreover, $\alpha' = m \le r' \le n$ if $I \cap F = \emptyset$, otherwise $\alpha' = r' \le n$ .  (Intuitively, if $\varepsilon$ is accepted by $\aut_{e'}$, then the value of $r'$ can be less than $m$.) Finally, $\aut_{(e')^{\{m, \infty\}}}$ is constructed by adapting $\alpha'$ in $\aut_{(e')^{\{m,n\}}}$ as follows: $\alpha' = m \le r'$ if $I \cap F = \emptyset$ and $\alpha' = \ltrue$ otherwise. 
%

@@

\item For each maximal sub-expression $e'$ of $e$ such that $e'$ contains no occurrences of counting operators, an NFA $\aut_{e'}$ can be constructed by structural induction on the syntax of $e'$. 
%Without loss of generality,  we assume that $R_{\aut_{e'_1}} \cap R_{\aut_{e'_2}} = \emptyset$. 
Then we have $\aut_{e'} = \aut_{e'_1} \concat \aut_{e'_2}$, $\aut_{e'} = \aut_{e'_1} \cup \aut_{e'_2}$, or $\aut_{e'} = \aut_{e'_1} \cap \aut_{e'_2}$. 
\end{enumerate}

%\begin{example}
%an example here
%\end{example}

@@

%%%%%%%%%%% original texts by denghang %%%%%
%%%%%%%%%%% original texts by denghang %%%%%
\hide{
The most important point of our approach is encoding counting operators by registers rather than explicitly unfolding them. In this section, we discuss the encoding in detail. A counting operator is called \emph{handled} if it is not the sub-regex of complement, closure, or another counting, and called \emph{unhandled} otherwise. For example, $(\Sigma \setminus a)^{\{m,n\}}$ is handled, while $((\Sigma \setminus a)^{\{m,n\}})^*$ is unhandled. We first discuss the encoding of handled counting operators and then the encoding of unhandled counting operators.

% We must syntactically rewrite bounded repetition $\regex\{m,n\}$ if it is the sub-regex of complement (e.g., $(\regex\{m,n\})^C$), closure(e.g., $(\regex\{m,n\})^*$), and bounded repetition(e.g., $(\regex\{m,n\})\{m', n'\}$), we unwind it to $\regex^m\mid\cdots\mid\regex^n$. After this syntactic rewriting, we call the resulting regex $\regex'$ \emph{non-nested}.
Given a regex $\regex$ without an unhandled counting operator, the base case is constructing the CEFA of a single character, an empty string, or an empty language. The inductive step is constructing CEFA of the concatenation, union, intersection, closure, complement, and counting of CEFA(s) of sub-regex(es). Operations except for counting are trivial and represented in Appendix \ref{appendix:cefa}; we only illustrate counting here.

Given a regex $\regex^{\{m,n\}}$ without unhandled counting operator, suppose the NFA recognizing $\regex$ is $\aut = (Q,\Sigma,\delta,q_I,F)$, then the CEFA recognizing $\regex^{\{m,n\}}$ is defined as $\aut_{m,n} = (Q, \Sigma, \delta',q_I, F\cup\{q_I\}, r, \theta)$ where:
\begin{itemize}
  \item $q_I$ is set to accepting state,
  \item $r$ is a new register,
  \item $\delta'$ is composed by transition $q\xrightarrow[(0)]{a} q'$ for each transition $q\xrightarrow[()]{a} q' \in \delta$, and transition $q_f\xrightarrow[(1)]{\epsilon} q_I$ for each accepting state $q_f\in F$,
  \item $\theta$ is the function mapping each accepting state to the same linear integer arithmetic $(m\leq )r\leq n$. The lower bound $m\leq r$ is unnecessary if $q_I\in F$.
\end{itemize}
The main idea is to add an epsilon transition to the initial state from each accepting state, like closure, to repeat the sub-regex. Then the counting time is incremented by 1 when the epsilon transition is invoked. Details are illustrated in Fig.\ref{fig:construct_repetition}. A new register $r$ is added to store the counting times. The transition $q\xrightarrow[(1)]{\epsilon} q_I$ is constructed to increment the counting times. The accepting condition $(m\leq) r\leq n$ is attached to each accepting state to constrain the counting times of the accepting word. After the construction, we need to eliminate epsilon transitions. It is done by adding transition $q_f\xrightarrow[(1)]{a} q$ for transition $q_I\xrightarrow[(0)]{a} q$ and then deleting the transition $q_f\xrightarrow[(1)]{\epsilon} q_I$, for each accepting state $q_f$.

\begin{figure}[h]
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \begin{tikzpicture}[
      shorten >=1pt,node distance=2cm,on grid,>={Stealth[round]},
      initial text=, every state/.style={minimum size = 0.001cm},
      ]

      \node[state,initial]            (q_0)                      {};
      \node[state, accepting]         (q_1) [right=of q_0]       {};
      \node [fit=(q_0) (q_1)] {$\cdots$};
    \end{tikzpicture}
    \caption{The NFA recognizing $\regex$}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \begin{tikzpicture}[
      shorten >=1pt,node distance=2cm,on grid,>={Stealth[round]},
      initial text=, every state/.style={minimum size = 0.001cm},
      accepting text=
      ]

      \node at (-0.5,0.5) {$\scriptstyle (\myemph{m}\leq) r\leq \myemph{n}$};
      \node[state,initial,accepting]            (q_0)                      {};
      \node[state, accepting]         (q_1) [right=of q_0]       {};
      \path[->] (q_1) edge[bend right]   node    [above] {\myemph{$\scriptstyle \mathbf{\epsilon:(1)}$}} (q_0);
      \node [fit=(q_0) (q_1)] {$\cdots$};
    \end{tikzpicture}
    \caption{The CEFA recognizing $\regex^{\{m,n\}}$}
  \end{subfigure}
  \caption{The construction of counting operator}
  \label{fig:construct_repetition}
\end{figure}
\begin{example}
  To construct the CEFA recognizing $(\Sigma\setminus a)\{5,60\}$ with alphabet $\Sigma = \{a,b,c\}$. As shown in Fig.\ref{subfig:count_aut_sigma_minus_a_5_60}, the base cases are the constructions of character $b$ (Fig.\ref{subfig:count_aut_b}) and $c$ (Fig.\ref{subfig:count_aut_c}). The first inductive step is building their union (Fig.\ref{sub@subfig:count_aut_sigma_minus_a}). The second inductive step is to add counting information on the automaton. To do that we add a new register $r$ to save counting times,  transitions $q_2\xrightarrow[(1)]{\epsilon} q_0$ and $q_1\xrightarrow[(1)]{\epsilon} q_0$  to update the counting times. The accepting condition of each accepting state is $5\leq r\leq 60$, which ensures the accepting word's counting time is in the range $[5,60]$. After epsilon elimination we get the CEFA recognizing $(\Sigma\setminus a)\{5,60\}$ (shown in Fig.\ref{subfig:count_aut_sigma_minus_a_5_60}).
  \begin{figure}[h]
    \begin{subfigure}[b]{0.49\textwidth}
      \centering
      \begin{tikzpicture}[
        shorten >=1pt,node distance=2cm,on grid,>={Stealth[round]},
        initial text=, every state/.style={minimum size = 0.001cm},
        ]
        \node[state,initial]            (q_0)                      {$\scriptstyle q_0$};
        \node[state,accepting]                    (q_1) [right=of q_0]       {$\scriptstyle q_1$};

        \path[->] (q_0) edge              node      [above]           {$\scriptstyle b:()$} (q_1);
      \end{tikzpicture}
      \caption{The NFA recognizing $b$}
      \label{subfig:count_aut_b}
    \end{subfigure}
    \begin{subfigure}[b]{0.49\textwidth}
      \centering
      \begin{tikzpicture}[
        shorten >=1pt,node distance=2cm,on grid,>={Stealth[round]},
        initial text=, every state/.style={minimum size = 0.001cm},
        ]
        \node[state,initial]            (q_0)                      {$\scriptstyle q_0$};
        \node[state,accepting]                    (q_1) [right=of q_0]       {$\scriptstyle q_2$};

        \path[->] (q_0) edge              node      [above]           {$\scriptstyle c:()$} (q_1);
      \end{tikzpicture}
      \caption{The NFA recognizing $c$}
      \label{subfig:count_aut_c}
    \end{subfigure}
    \begin{subfigure}[b]{0.49\textwidth}
      \centering
      \begin{tikzpicture}[
        shorten >=1pt,node distance=1cm,on grid,>={Stealth[round]},
        initial text=, every state/.style={minimum size = 0.001cm},
        ]
        \node[state,initial]      (q_0)                      {$\scriptstyle q_0$};
        \node (q_tmp) [right=of q_0]       {};
        \node[state, accepting]   (q_1) [above=of q_tmp]       {$\scriptstyle q_1$};
        \node[state, accepting]   (q_2) [below=of q_tmp]       {$\scriptstyle q_2$};

        \path[->] (q_0) edge node [right] {$\scriptstyle b:()$} (q_1)
        (q_0) edge           node [right] {$\scriptstyle c:()$} (q_2);
      \end{tikzpicture}
      \caption{The NFA recognizing $\Sigma \setminus a$}
      \label{subfig:count_aut_sigma_minus_a}
    \end{subfigure}
    \begin{subfigure}[b]{0.49\textwidth}
      \centering
      \begin{tikzpicture}[
        shorten >=1pt,node distance=1cm,on grid,>={Stealth[round]},
        initial text=, every state/.style={minimum size = 0.005cm},
        ]
        \node at (-1, 1) {$\scriptstyle \myemph{5}\leq r \leq \myemph{60}$};
        \node[state,initial, accepting]      (q_0)                      {$\scriptstyle q_0$};
        \node (q_tmp) [right=of q_0]       {};
        \node[state, accepting]   (q_1) [above=of q_tmp]       {$\scriptstyle q_1$};
        \node[state, accepting]   (q_2) [below=of q_tmp]       {$\scriptstyle q_2$};

        \path[->] (q_0) edge node [right] {$\scriptstyle b:(0)$} (q_1)
        (q_1) edge [loop right] node {$\scriptstyle b:(\myemph{1})$} (q_1)
        (q_0) edge           node [right] {$\scriptstyle c:(0)$} (q_2)
        (q_2) edge [loop right] node {$\scriptstyle c:(\myemph{1})$} (q_0);
      \end{tikzpicture}
      \caption{The CEFA recognizing $(\Sigma\setminus a)\{\myemph{5},\myemph{60}\}$}
      \label{subfig:count_aut_sigma_minus_a_5_60}
    \end{subfigure}
    \caption{The example of constructing regex with counting}
  \end{figure}
\end{example}

Above discusses the construction of handled counting operators. In the same way, we can not construct the unhandled counting operator. For example, for the regex $(ab^{\{1,100\}})*$ with unhandled counting operator, there are infinite repetitions of $ab^{\{1,100\}}$. The counting times of each repetition are irrelevant, so we can not represent them using finite registers. We must syntactically rewrite the unhandled counting operator into union and concatenation operators. More exactly, the sub-regex $\regex^{\{m,n\}}$ in the unhandled regex is rewritten into $\underbrace{\regex\cdot\regex\cdots\regex}_{m \text{ times}}\mid\cdots\mid \underbrace{\regex\cdot\regex\cdots\regex}_{n \text{ times}}$.
% Based on the decision procedure defined in the paper \cite{atva2020}, we propose an efficient algorithm to solve ESL formula $\varphi\wedge \psi$ where $\varphi$ is the conjunction of literals and $\psi$ is the conjunction of linear literals without length operation. For each string variable $x$, we first compute pre-images of all linear literal $i=|x|$. Then we generate CEFAs of all regular expressions $\regex$ in all regular literals $x\in \regex$ and intersect these CEFAs and pre-images to get the final automaton $\aut_x$ for $x$. There is a final CEFA for each string variable $x$. The linear arithmetic constraint $\psi$ restricts the accepting word of final CEFAs. Thus the satisfiability problem is the emptiness checking problem of final CEFAs under $\psi$. If it is empty, the string constraints are unsatisfiable. Otherwise, the string constraints are satisfiable.
% The emptiness checking problem of CEFAs under linear integer arithmetic is theoretically pspace-complete\cite{atva2020}. To solve it efficiently for a practical example, we develop heuristic ways such as under-approximation (Section \ref{subsec:under-approx}) and symbolic-aware simplification (Section \ref{subsec:simplify}).

%\subsection{Handling of String-length}
Given a string-length constraint, we build a CEFA with a register storing the word length. Namely, the CEFA is $(\{q_I\}, \Sigma, \{q_I\xrightarrow[(1)]{\Sigma} q_I\}, q_I, \{q_I\}, r, \theta)$ where accpeting condition $\theta$ is based on the length formula. As an illustration, example \ref{eg:pre_len} shows the CEFA of length constraints $|x| > 120$, whose language is the set of all words with a length greater than $120$. The simple loop $q_I\xrightarrow[(1)]{\Sigma} q_I$ increase $1$ on register $r$ for all char in $\Sigma$, so that the value of $r$ is equal to the length of the accepting word. The accepting condition $\theta$ attaches $q_I$ to $r > 120$, ensuring the length of the accepting word is greater than $120$.

% The formal definition of pre-images of abundant string functions is listed in \cite{atva2020}, and we do not repeat it here.
\begin{example}[The CEFA of $|x| > 120$]\label{eg:pre_len} 
  \begin{figure}[h]
    \centering
    \begin{tikzpicture}[
      shorten >=1pt,node distance=2cm,on grid,>={Stealth[round]},
      initial text=, every state/.style={minimum size = 0.001cm},
      ]

      \node at (-0.5, 0.5) {$\scriptstyle r > 120$};
      \node[state,initial,accepting]            (q_0)       {$\scriptstyle q_I$};

      \path[->] (q_0) edge [loop below] node{$\scriptstyle\Sigma:(1)$} ();
    \end{tikzpicture}
  \end{figure}
\end{example} 
% \subsection{Combine Regex-counting and String-length}
% In the previous section, we discussed the construction of the regex-counting operator and string-length constraint. In this section, we show how to combine them. 
% \begin{definition} [The Product of Two CEFA]

%   Given two CEFAs $\aut_1 = (Q_1, \Sigma, \delta_1, q_{I_1}, F_1, R_1, \theta_1)$ and $\aut_2 = (Q_2, \Sigma, \delta_2, q_{I_2}, F_2, R_2, \theta_2)$ with $R_1\cap R_2 = \emptyset$, the product of them, denoted by $\aut_1 \times \aut_2$, is defined as $(Q_1\times Q_2, \Sigma, \delta, (q_{I_1}, q_{I_2}), F_1\times F_2, R_1\cup R_2, \theta)$ where
%   \begin{itemize}
%     \item $\delta$ comprises the tuples $((q_1, q_2), a, (q_1', q_2'), \myvec{v_1}\cdot \myvec{v_2})$ such that $(q_1, a, q_1', \myvec{v_1})\in \delta_1$ and $(q_2, a, q_2', \myvec{v_2})\in \delta_2$.
%     \item $\theta((q_1, q_2))$ is $\theta_1(q_1)\wedge \theta_2(q_2)$ for each accepting state $(q_1, q_2)\in F_1\times F_2$.
%   \end{itemize}
% \end{definition}
% With the help of product, we can combine the CEFAs of regex-counting and string-length. For example, the CEFA of $x$ in the string constraints $x \in (\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}} \wedge x \in \Sigma^* c \wedge |x| > 120$ is the product of the CEFA of regex $(\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}}$, the CEFA of regex $\Sigma^*$ and the CEFA of length constraints $|x| > 120$.
}
%%%%%%%%%%% original texts by denghang %%%%%
%%%%%%%%%%% original texts by denghang %%%%%

\vspace{-2mm}
\subsection{Solving $\cefadec$} \label{subsec:cefadec}
\vspace{-1mm}

@@

%At first, we compute CEFA $\cB_{i} = \aut_{i,1} \cap \cdots \cap \aut_{i, l_i}$ for each $i \in [n]$. For each $i \in [n]$, let $\Lambda'_{x_i} = \{\cB_i\}$. Evidently, $R_{\cB_i}= \bigcup \limits_{j \in [l_i]} R_{\aut_{i, j}}$ for each $i \in [n]$.
After Step 1, the nonemptiness of CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ w.r.t. $\varphi$ is reduced to the nonemptiness of CEFA in $\Lambda'_{x_1}, \cdots, \Lambda'_{x_n}$ w.r.t. $\varphi$.

%CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ are nonempty w.r.t. $\varphi$ 
%

%Therefore, to solve the nonemptiness of CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ w.r.t. $\varphi$, it is sufficient to compute $\cefaout(\cB_1)$

%Since the construction of $\cB_1, \cdots, \cB_n$ involves the computation of the product of multiple CEFA, it incurs an exponential blow-up in the worst case. Therefore, it is vital to reduce the sizes of these CEFAs. In the sequel, we propose some techniques to reduce the automata sizes.

@@

\vspace{-2mm}
\begin{observation}\label{obs-cefa-output}
CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ are nonempty w.r.t. $\varphi$ iff there are $\myvec{v_{1}} \in \cefaout(\cB_1), \cdots, \myvec{v_{n}} \in \cefaout(\cB_n)$ such that $\varphi[\myvec{v_{1}}/R_{\cB_{1}}, \cdots, \myvec{v_{n}}/R_{\cB_{n}}]$ is true.
\end{observation}
\vspace{-1mm}

%From the fact that $\Lambda'_{x_i}$ for each $i \in [n]$ contains exactly one CEFA, we deduce that the following two conditions are equivalent.  
%\begin{itemize}
%\item CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ are nonempty w.r.t. $\varphi$, namely, 
%there are an assignment $\theta: \{x_1, \cdots, x_n\} \rightarrow \Sigma^*$ and vectors $\myvec{v_{1}}, \cdots, \myvec{v_{n}}$ such that $(\theta(x_1); \myvec{v_1}) \in \Lang(\cB_1)$, $\cdots$, $(\theta(x_n); \myvec{v_n}) \in \Lang(\cB_n)$ and $\varphi[\myvec{v_{1}}/R_{\cB_{1}}, \cdots, \myvec{v_{n}}/R_{\cB_{n}}]$ is true,
%
%\item there are $\myvec{v_{1}} \in \cefaout(\cB_1), \cdots, \myvec{v_{n}} \in \cefaout(\cB_n)$ such that $\varphi[\myvec{v_{1}}/R_{\cB_{1}}, \cdots, \myvec{v_{n}}/R_{\cB_{n}}]$ is true.
%\end{itemize}

@@

\vspace{-2mm}
\begin{observation}\label{obs-cefa-nfa}
For each CEFA $\aut = (R, Q, \Sigma, \delta, I, F, \alpha)$, 
%
$$\cefaout(\aut) = \left\{\sum \limits_{\myvec{v} \in \costset_\aut} \eta(\myvec{v}) \myvec{v} \  \big{|}\ \eta \in \parikh(\cU_\aut) \mbox{ and }  \alpha\left[ \sum \limits_{\myvec{v} \in \costset_\aut} \eta(\myvec{v}) \myvec{v} \big{/} R \right] \mbox{ is true}\right\}.$$
%
\end{observation}
\vspace{-1mm}

For $i \in [n]$, let $\alpha_i$ be the accepting condition of $\cB_i$. 
Then from Observation~\ref{obs-cefa-nfa}, we know that the following two conditions are equivalent, 
\begin{itemize}
%\item CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ are nonempty w.r.t. $\varphi$, namely, 
%there are an assignment $\theta: \{x_1, \cdots, x_n\} \rightarrow \Sigma^*$ and vectors $\myvec{v_{1}}, \cdots, \myvec{v_{n}}$ such that $(\theta(x_1); \myvec{v_1}) \in \Lang(\cB_1)$, $\cdots$, $(\theta(x_n); \myvec{v_n}) \in \Lang(\cB_n)$ and $\varphi[\myvec{v_{1}}/R_{\cB_{1}}, \cdots, \myvec{v_{n}}/R_{\cB_{n}}]$ is true,
%
\item there are $\myvec{v_{1}} \in \cefaout(\cB_1), \cdots, \myvec{v_{n}} \in \cefaout(\cB_n)$ such that $\varphi[\myvec{v_{1}}/R_{\cB_{1}}, \cdots, \myvec{v_{n}}/R_{\cB_{n}}]$ is true, 
%CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ are nonempty w.r.t. $\varphi$ 
%
\item 
there are $\eta_1 \in \parikh(\cU_{\cB_1}), \cdots, \eta_n \in \parikh(\cU_{\cB_n})$ such that 
%
%\begin{equation}\label{eqn-cefa-lia}
\[
\bigwedge \limits_{i \in [n]} \alpha_i\left[ \sum \limits_{\myvec{v} \in \costset_{\cB_i}} \eta_i(\myvec{v}) \myvec{v} \big{/}R_{\cB_i}  \right] \wedge \varphi \left[\sum \limits_{\myvec{v} \in \costset_{\cB_1}} \eta_1(\myvec{v}) \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum \limits_{\myvec{v} \in \costset_{\cB_n}} \eta_n(\myvec{v}) \myvec{v} \big{/}R_{\cB_n} \right]
\]
%\end{equation} 
%
is true. 
\end{itemize}

>>>>

%!TEX root = ../main.tex

@@

\begin{definition}[Nonemptiness of CEFA w.r.t. an LIA formula
, abbreviated as $\cefadec$]
Suppose that $x_1, \cdots, x_n$ are string variables, $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are nonempty sets of CEFA over the alphabet $\Sigma$ with $\Lambda_{x_i} = \{\aut_{i, 1}, \cdots, \aut_{i, l_i}\}$ for every $i \in [n]$, and the sets of registers $R_{\aut_{1, 1}}, \cdots, R_{\aut_{1, l_1}}, \cdots, R_{\aut_{n, 1}}, \cdots, R_{\aut_{n, l_n}}$ are mutually disjoint, moreover, $\varphi$ is an LIA formula whose free variables are from $ \bigcup \limits_{i \in [n], j \in [l_i]} R_{\aut_{i, j}}$. Then  the CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are said to be nonempty w.r.t. $\varphi$ if there are assignments $\theta: \{x_1, \cdots, x_n\} \rightarrow \Sigma^*$ and vectors $\myvec{v_{i,j}}$
such that $(\theta(x_i); \myvec{v_{i,j}}) \in \Lang(\aut_{i, j})$ and  $\varphi[(\myvec{v_{i,j}}/R_{\aut_{i,j}})]$ is true, for every $i \in [n], j \in [l_i]$.
\end{definition}
\begin{proposition}\label{prop-cefadec}
$\cefadec$ is PSPACE-complete. 
\end{proposition}
\begin{proof}
  The lower bound follows from the fact that the nonemptiness problem of the intersection of NFAs is PSPACE-complete \cite{pspace_int}. For the upper bound, \cite{atva2020} showed that $\cefadec$ is in PSPACE by reducing it to monotonic counter machines.
\end{proof}
Note that the decision procedure in \cite{atva2020} was only used to prove the upper bound in Proposition~\ref{prop-cefadec} and not implemented as a matter of fact. Instead, the symbolic model checker nuXmv was used to solve $\cefadec$. We do not rely on nuXmv in this work and shall propose a new algorithm for solving $\cefadec$ in Section~\ref{subsec:cefadec}.

\subsection{From $\reclsat$ to $\cefadec$} \label{subsec:regex2cefa}

@@

Let $e$ be a register-representable regular expression over $\Sigma$. By the following procedure, we will construct a CEFA out of $e$, denoted by $\aut_e$. 
\begin{enumerate}
%
\item For each sub-expression $(e')^{\{m,n\}}$ with $m \le n$ (resp. $(e')^{\{m, \infty\}}$) of $e$, we construct a CEFA $\aut_{(e')^{\{m,n\}}}$ (resp. $\aut_{(e')^{\{m, \infty\}}}$) as follows. Let $\aut_{e'} = (Q, \Sigma, \delta, I, F)$. Same as the construction for Kleene star except for the new register and updates on transitions, $\aut_{(e')^{\{m,n\}}} = ((r'), Q', \Sigma, \delta'', I', F', \alpha')$, where $r'$ is a new register, 
%
$$
\begin{array}{l c l}
\delta'' & = & \{(q, a, q', (0)) \mid (q, a, q') \in \delta\}\  \cup \\
& & \{(q_0, a, q', (1))  \mid \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\}\ \cup \\
& & \{(q, a, q', (1)) \mid q \in F, \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\},
\end{array}
$$ 
%
moreover, $\alpha' = m \le r' \le n$ if $I \cap F = \emptyset$, otherwise $\alpha' = r' \le n$ .  (Intuitively, if $\varepsilon$ is accepted by $\aut_{e'}$, then the value of $r'$ can be less than $m$.) Finally, $\aut_{(e')^{\{m, \infty\}}}$ is constructed by adapting $\alpha'$ in $\aut_{(e')^{\{m,n\}}}$ as follows: $\alpha' = m \le r'$ if $I \cap F = \emptyset$ and $\alpha' = \ltrue$ otherwise. 
%

@@

\item For each maximal sub-expression $e'$ of $e$ such that $e'$ contains no occurrences of counting operators, an NFA $\aut_{e'}$ can be constructed by structural induction on the syntax of $e'$. 
Then we have $\aut_{e'} = \aut_{e'_1} \concat \aut_{e'_2}$, $\aut_{e'} = \aut_{e'_1} \cup \aut_{e'_2}$, or $\aut_{e'} = \aut_{e'_1} \cap \aut_{e'_2}$. 
\end{enumerate}

@@

\subsection{Solving $\cefadec$} \label{subsec:cefadec}

@@

After Step 1, the nonemptiness of CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ w.r.t. $\varphi$ is reduced to the nonemptiness of CEFA in $\Lambda'_{x_1}, \cdots, \Lambda'_{x_n}$ w.r.t. $\varphi$.

@@

\begin{observation}\label{obs-cefa-output}
CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ are nonempty w.r.t. $\varphi$ iff there are $\myvec{v_{1}} \in \cefaout(\cB_1), \cdots, \myvec{v_{n}} \in \cefaout(\cB_n)$ such that $\varphi[\myvec{v_{1}}/R_{\cB_{1}}, \cdots, \myvec{v_{n}}/R_{\cB_{n}}]$ is true.
\end{observation}

@@

\begin{observation}\label{obs-cefa-nfa}
For each CEFA $\aut = (R, Q, \Sigma, \delta, I, F, \alpha)$, 
%
$$\cefaout(\aut) = \left\{\sum \limits_{\myvec{v} \in \costset_\aut} \eta(\myvec{v}) \myvec{v} \  \big{|}\ \eta \in \parikh(\cU_\aut) \mbox{ and }  \alpha\left[ \sum \limits_{\myvec{v} \in \costset_\aut} \eta(\myvec{v}) \myvec{v} \big{/} R \right] \mbox{ is true}\right\}.$$
%
\end{observation}

For $i \in [n]$, let $\alpha_i$ be the accepting condition of $\cB_i$. 
Then from Observation~\ref{obs-cefa-nfa}, we know that the following two conditions are equivalent, 
\begin{itemize}

\item there are $\myvec{v_{1}} \in \cefaout(\cB_1), \cdots, \myvec{v_{n}} \in \cefaout(\cB_n)$ such that $\varphi[\myvec{v_{1}}/R_{\cB_{1}}, \cdots, \myvec{v_{n}}/R_{\cB_{n}}]$ is true, 
\item 
there are $\eta_1 \in \parikh(\cU_{\cB_1}), \cdots, \eta_n \in \parikh(\cU_{\cB_n})$ such that 
\[
\bigwedge \limits_{i \in [n]} \alpha_i\left[ \sum \limits_{\myvec{v} \in \costset_{\cB_i}} \eta_i(\myvec{v}) \myvec{v} \big{/}R_{\cB_i}  \right] \wedge \varphi \left[\sum \limits_{\myvec{v} \in \costset_{\cB_1}} \eta_1(\myvec{v}) \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum \limits_{\myvec{v} \in \costset_{\cB_n}} \eta_n(\myvec{v}) \myvec{v} \big{/}R_{\cB_n} \right]
\]
is true. 
\end{itemize}
